---
title: "RNA-seq Report"
author: "Alejandro Reyes, asr2878"
date: "11/9/2021"
output: html_document
---

This project was conducted to understand how the differences in gene expression 
between time of collection vary by genotypes of *Arabidopsis thaliana*. This code 
uses DESeq to identify genes with significant interaction terms between time and
genotype. Here is the given descriptions for the genotypes of the plant 
*Arabidopsis thaliana*:
1. Col wild-type *Arabidopsis thaliana*.
2. 14B a mutant lacking two genes for a plant specific translation initiation 
   factor, eIFiso4G1 (AT5g57870) and eIFiso4G2 (At2g24050).
3. 4G a mutant lacking the one gene for eIF4G (AT3g60240), a translation factor 
   that is more  similar to those of other eukaryotes.

```{r message=FALSE}
source("bio321g_rnaseq_utils.R")
library(BiocParallel)
register(SnowParam(4))
library(DESeq2)
library(ggplot2)
library(pheatmap)
```

# 1. 
Initially a DESeqDataSeq object, "dds" was created to test the significance, if 
any, of the interaction term time:genotype from the genes that had sufficient 
expression levels. 1429 genes had a significant interaction term (FDR ≤ 0.10). 
Of these genes, 142.9 are expected to be false discoveries. This was determined 
by multiplying the number of expected genes (1429) by the false discovery rate 
(FDR = .10).
```{r message=FALSE}
dds = DESeqDataSetFromMatrix(
  countData = rnaCounts,
  colData = sampleAnnotation,
  design = ~ time + genotype + time:genotype)

dds = DESeq(dds, test = "LRT", reduced = ~ time + genotype, parallel=TRUE)
interaction_results = results(dds)
interaction_results = interaction_results[order(interaction_results$pvalue),]

testedGenes = which(!is.na(interaction_results$padj))  # actual num of sig tests
# NA implies the gene expression was too low

FDR = .1  # control on padj
numGenesSigIR = sum(interaction_results[testedGenes, "padj"] <= FDR)
numExpectedFalseDiscoveries = numGenesSigIR * FDR
```


# 2.
The normalized counts from dds, after the data was normalized with the DESeq 
function, were then extracted. These counts were then filtered by removing genes
without sufficient data. The data was log2 transformed with an offset of 1 to 
reduce heteroskedacity.
```{r}
normed = counts(dds, normalized=TRUE)
normed = normed[rowSums(normed) > 0,] # remove genes with no data
lgNorm = log2(normed + 1) # log transform; satisfies heteroskedacity
```

# 3.
A PCA plot was then constructed using the transformed data. Visually, the 
samples collected in the dark were most separated from the samples collected in
the light along the PC1 direction (dark was oriented in the positive side of the
plot while light was negative). Using a similar visual analysis, the Col and 4G 
genotypes of *Arabidopsis thaliana* were most separated from the 14B genotype 
along the PC2 direction.
```{r}
pca_lgNorm = prcomp(t(lgNorm))

pca_lgNorm_annotation = data.frame(
  row.names = colnames(normed),
  group = gsub("^X|\\d$", "", colnames(normed))
)

## set up data.frame pcaData for ggplot...
pcaData = data.frame(pca_lgNorm$x[ , 1:2])
## add in sample annotation info
pcaData$group = pca_lgNorm_annotation[rownames(pcaData), "group"]
## and sample names
pcaData$sample = rownames(pcaData)

gg = ggplot(pcaData, aes(x=PC1, y=PC2, color=group, label=sample))
gg = gg + geom_point(size=2.5, alpha=0.75)
gg = gg + scale_color_manual(values=groupColors)
print(gg)
```

# 4. 
A vector was created containing the assigned genes (gene_ontology_primary_id 
value of GO:0071495). From the assigned genes, the gene ID, gene symbol, and 
gene description were complied in a tsv file (attached and) named 
assigned_gene_set.tsv which has been provided.
```{r}
# get assigned gene set
target_set = goAssociations[goAssociations$gene_ontology_primary_id 
                            == "GO:0071495",]
# length of 82

# df with gene id, symbol, and description
assigned_genes = geneNamesAndDescriptions[
                    is.element(geneNamesAndDescriptions$gene, target_set$gene),]
row.names(assigned_genes) = NULL

# write .tsv file
write.table(assigned_genes, file="assigned_gene_set.tsv", 
            quote=FALSE, row.names = FALSE, sep="\t")
```


# 5. 
The normalized (and transformed) data was further refined to only include the 
genes assigned.
```{r}
lgGO = lgNorm[is.element(row.names(lgNorm), target_set$gene),]
```

# 6.
A principal components analysis was then preformed on this data (lgGO). Similar 
to the previous PCA plot, the  samples were separated based on time of 
collection along the PC1 axis (the dark samples were centered around 5 while the
day samples were centered around -6). Interestingly, the separation of the 
genotypes were inverted from the previous full gene set PCA plot; the 14B 
genotypes were more negative (~-3) along the PC2 axis while the Col and 4G 
genotypes were more positive (~2 with an outlier at ~4).
```{r}
pca_lgGO = prcomp(t(lgGO))
pcaDatalgGO = data.frame(pca_lgGO$x[ , 1:2])
pcaDatalgGO$group = pca_lgNorm_annotation[rownames(pcaDatalgGO), "group"]
pcaDatalgGO$sample = rownames(pcaDatalgGO)
gg2 = ggplot(pcaDatalgGO, aes(x=PC1, y=PC2, color=group, label=sample))
gg2 = gg2 + geom_point(size=2.5, alpha=0.75)
gg2 = gg2 + scale_color_manual(values=groupColors)
print(gg2)
```

# 7.
The differences in the log-transformed normalized expression level and the mean 
expression level of each of these genes were stored in heatData. A clustered 
heatmap was then constructed using this data and the pheatmap function. The 
expression level in heatData was limited by setting values greater than 2 equal 
to 2 and values less than -2 equal to -2. This was done to make the resulting 
pheatmap more readable and to prevent the overabundance of a few individual 
colors.

The two largest clusters were found in the upper left and right corners of the 
heatmap. The genes in the upper left corner of the heatmap have mostly low 
expression. It is interesting to note that these samples were all collected in 
the light and were independent of the genotype. The genes in the upper right 
corner of the heatmap have an above average rate of expression. These samples 
were also independent of genotype but were dependent on the time of collection 
being dark. The LHCB2.2 gene was highly expressed when the sample was collected 
in the dark and inhibited when collected in the light. There were also two more 
clusters that formed on the lower corners of the heatmap. They showed an inverse
relationship of expression and time of collection than that of the two larger 
clusters mentioned previously. 
```{r fig.asp = 1.25}
heatData = lgGO - rowMeans(lgGO)
heatData[heatData > 2] = 2; heatData[heatData < -2] = -2
# help("pheatmap")
pheatmap(
  heatData,
  heatPalette,
  clustering_method = "average",
  # show_rownames = FALSE, 
  labels_row=geneNamesAndDescriptions[rownames(heatData), "symbol"],
  cellheight = 5,
  cellwidth = 20,
  fontsize_row = 5,
  fontsize_col = 5,
  angle_col = 90
)
```

# 8.
Lastly, the lgGO data was filtered by the lowest interaction term 
(time:genotype) according to pvalues in the DESeq results of the dds object. 
The 9 genes with this lowest interaction term were then used to construct a 
stripchart with the provided stripchart321g function. 

The LHCB2.2, CKA4, bHLH129, KNAT5, PUB9, and MAPKKK5 genes generally had greater
levels of expression in the samples collected in the dark than in the light 
across all genotypes (MAPKKK5 had an overlap in genotypes 4G and COL while PUB9 
had an overlap in 14B and 4G). The IRT1, UF3GT, and SAG29 genes had cases where 
this relationship was inverted: in the 14B and 4G genotypes of SAG29, the 14B 
genotype of IRT1, and across all genotypes of UF3GT. The most extreme cases of 
differences in expression between timepoints that vary across genotypes were 
seen in the genes CKA4, PUB9, SAG9, UF3GT, and IRT1. In these genes, two of the
genotypes had similar expression levels while one was visually different. In the
case of SAG29 and PUB9, the outlying genotype was COL; the light samples were 
less expressed while the dark samples were more expressed than the other two 
genotypes. In CKA4 and UF3GT, the 14B genotype was significantly different from 
the other two. In both genes, the light samples were more expressed. In the IRT1 
gene, the light sample of the 4G genotype was more expressed than its 
counterparts in the other two genotypes. KNAT5 and LHCB2.2 had relatively 
similar expression levels across genotypes except for one light sample being 
expressed less than the others (the 14B genotype of LHCB2.2 and the COL genotype
of KNAT5). In MAPKKK5, dark samples were similarly expressed while the light 
samples of the three genotypes had no similarity between each other; the 14B 
genotype had the lowest expression levels, 4G had the second lowest, and COL had
the highest. It would be best to describe the mean of expression levels and the 
margin of the differential expression between timepoints since there was next to
no overlap in the samples of the bHLH129 gene. The mean expression, in ascending
order by genotype, is COL, 14B, and 4G while the greatest margin of differential
expression between timepoints, in ascending order by genotype, is 14B, COL, and 
4G.
```{r message=FALSE}
# use previous deseq results but only include assigned genes
# top 9 genes with the smallest p values
reduced_DESeq = interaction_results[is.element(row.names(interaction_results), 
                                              target_set$gene), ]
reduced_DESeq = reduced_DESeq[order(reduced_DESeq$pvalue), ]
reduced_DESeq = reduced_DESeq[1:9, ]
reduced_lgGO = lgGO[is.element(row.names(lgGO), row.names(reduced_DESeq)), ]
stripchart_rlgGO = stripchart321g(
                                  reduced_lgGO,
                                  sampleAnnotation,
                                  geneNames = geneNamesAndDescriptions,
                                  colorValues = groupColors)
stripchart_rlgGO
```
